\ripple: Differentiable and Hardware-Accelerated Waveforms for Gravitational Wave Data Analysis

### 2023-02-07 src/tex/ms.tex
fix adam's ciela affiliation
typos, affiliation, preprint

\newcommand{\JHU}{\affiliation{William H. Miller III Department of Physics and Astronomy, Johns Hopkins University, Baltimore, Maryland 21218, USA}} 
\newcommand{\OKC}{\affiliation{The Oskar Klein Centre, Department of Physics, Stockholm University, AlbaNova, SE-106 91 Stockholm, Sweden}} 
\newcommand{\NORDITA}{\affiliation{Nordic Institute for Theoretical Physics (NORDITA), 106 91 Stockholm, Sweden}}
\newcommand{\Ciela}{\affiliation{Ciela -- Computation and Astrophysical Data Analysis Institute, Montréal, Quebec, Canada}}
\newcommand{\UdeM}{\affiliation{Département de Physique, Université de Montréal, 1375 Avenue Thérèse-Lavoie-Roux, Montréal, QC H2V 0B3, Canada}} 
\newcommand{\Mila}{\affiliation{Mila -- Quebec AI Institute, 6666 St-Urbain, \#200, Montreal, QC, H2S 3H1}} 
\newcommand{\CGP}{\affiliation{Center for Gravitational Physics, University of Texas at Austin, Austin, TX 78712, USA}}
\newcommand{\CCA}{\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}}
\newcommand{\CUHK}{\affiliation{Department of Physics, The Chinese University of Hong Kong, Shatin, N.T., Hong Kong}}

@@

\author{Adam Coogan} \UdeM \Mila
\author{Daniel Foreman-Mackey} \CCA
\author{Maximiliano Isi} \CCA
\author{Aaron Zimmerman} \CGP

@@

\begin{figure}[t]
    \script{random_matches.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/random_matches.pdf}
        \caption{
            Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
            % The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
            % The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
            Points which give $\mathcal{M}=0$ (up to numerical precision) are plotted as the lowest value in the colorbar
            It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
            % \te{Ok so some of the remaining error was coming from a weird thing at the cutoff of the waveform. I've now fixed this, so error is further down. Let me know if you still think it needs editing.}
            % \dfm{I think this figure could use some work. Maybe the points could be smaller to make them easier to resolve. And it could be interesting to clip the mismatch to something like $10^{-16}$ so that we don't lose the points with zero mismatch. Right now it looks less impressive!}
            % \mi{I agree. Furthermore, it looks like there's a subpopulation of darker points hiding under the brighter ones up to and around $M \sim 100$? is there a dependence on the mass ratio? it'd be good to comment on this, and maybe show a corner plot adding $q$ to the mix. Also, $M$ is lacking units (I assume $M_\odot$, but it should be shown in the label)}
        }
        \label{fig:match}
    \end{centering}
\end{figure}

@@

One of the primary s of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite-difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine-precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension added, $D$, one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
For a review of AD methods see~\cite{2018arXiv181105031M}.
% \kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond-GR models, waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases many more parameters can be added.

@@

\section{Acknowledgments}
This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation.
This work was supported by collaborative visits funded by the Cosmology and Astroparticle Student and Postdoc Exchange Network (CASPEN). 
T.E.\ is supported by the Horizon Postdoctoral Fellowship.
A.Z.~is supported by NSF Grant PHY-2207594.
A.~C.\ acknowledges funding from the Schmidt Futures Foundation.

>>>>

\newcommand{\JHU}{\affiliation{William H. Miller III Department of Physics and Astronomy, Johns Hopkins University, Baltimore, Maryland 21218, USA}} 
\newcommand{\OKC}{\affiliation{The Oskar Klein Centre, Department of Physics, Stockholm University, AlbaNova, SE-106 91 Stockholm, Sweden}} 
\newcommand{\NORDITA}{\affiliation{Nordic Institute for Theoretical Physics (NORDITA), 106 91 Stockholm, Sweden}}
\newcommand{\Ciela}{\affiliation{Ciela -- Computation and Astrophysical Data Analysis Institute, Montréal, Quebec, Canada}}
\newcommand{\UdeM}{\affiliation{Département de Physique, Université de Montréal, 1375 Avenue Thérèse-Lavoie-Roux, Montréal, QC H2V 0B3, Canada}} 
\newcommand{\Mila}{\affiliation{Mila -- Quebec AI Institute, 6666 St-Urbain, \#200, Montreal, QC, H2S 3H1}} 
\newcommand{\WI}{\affiliation{Weinberg Institute, University of Texas at Austin, Austin, TX 78712, USA}}
\newcommand{\CGP}{\affiliation{Center for Gravitational Physics, University of Texas at Austin, Austin, TX 78712, USA}}
\newcommand{\CCA}{\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}}
\newcommand{\CUHK}{\affiliation{Department of Physics, The Chinese University of Hong Kong, Shatin, N.T., Hong Kong}}

@@

\author{Adam Coogan} \Ciela \UdeM \Mila
\author{Daniel Foreman-Mackey} \CCA
\author{Maximiliano Isi} \CCA
\author{Aaron Zimmerman} \WI

@@

\begin{figure}[t]
    \script{random_matches.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/random_matches.pdf}
        \caption{
            Match between the \ripple and \lalsuite implementations of the IMRPhenomD waveform as a function of total mass and effective spin. 
            % The lower density of points at low masses is caused by the match being exactly 1.0 and therefore not appearing in the log scale. 
            % The gray points are where the match is exactly 1.0 and therefore would not appear in the log scale.
            Points which give $\mathcal{M}=0$ (up to numerical precision) are plotted as the lowest value in the colorbar.
            It is clear that the \ripple waveform matches the \lalsuite implementation many of orders of magnitude more than necessary for all data analysis tasks across the entire parameter space.
            % \te{Ok so some of the remaining error was coming from a weird thing at the cutoff of the waveform. I've now fixed this, so error is further down. Let me know if you still think it needs editing.}
            % \dfm{I think this figure could use some work. Maybe the points could be smaller to make them easier to resolve. And it could be interesting to clip the mismatch to something like $10^{-16}$ so that we don't lose the points with zero mismatch. Right now it looks less impressive!}
            % \mi{I agree. Furthermore, it looks like there's a subpopulation of darker points hiding under the brighter ones up to and around $M \sim 100$? is there a dependence on the mass ratio? it'd be good to comment on this, and maybe show a corner plot adding $q$ to the mix. Also, $M$ is lacking units (I assume $M_\odot$, but it should be shown in the label)}
        }
        \label{fig:match}
    \end{centering}
\end{figure}

@@

One of the primary aims of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite-difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine-precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension added, $D$, one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
For a review of AD methods see~\cite{2018arXiv181105031M}.
% \kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond-GR models, waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases many more parameters can be added.

@@

\section{Acknowledgments}
This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation.
This work was supported by collaborative visits funded by the Cosmology and Astroparticle Student and Postdoc Exchange Network (CASPEN). 
T.E.\ is supported by the Horizon Postdoctoral Fellowship.
A.Z.~is supported by NSF Grant PHY-2207594.
A.~C.\ acknowledges funding from the Schmidt Futures Foundation.
This document has been assigned preprint numbers UTWI-4-2023 and LIGO-P2300025.
